---
title: "Determining Factors for Delaying Medical Treatment"
author: "Nandita Kannapadi"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
subtitle: NHIS 2024 Multivariable Logistic Regression Analysis
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 4.5
)
```

```{r load-packages}
library(dplyr)
library(ggplot2)
library(broom)
library(readr)
```

## Research Question

Which factors — insurance coverage, race/ethnicity, or educational
attainment — are most strongly associated with delaying medical care due
to cost among U.S. adults? Using the 2024 National Health Interview
Survey (NHIS) Sample Adult file, this analysis fits bivariate and
multivariable logistic regression models to separate the independent
contribution of each factor.

## Reading the Data

```{r read-data}
# Data: NHIS 2024 Sample Adult file, downloaded from
# https://www.cdc.gov/nchs/nhis/documentation/2024-nhis.html
adult2024 <- read_csv("national_health_interview_survey.csv")

n_raw <- nrow(adult2024)
n_raw
```

## Cleaning: Variable-Specific Missing Value Handling

`HISPALLP_A` and `EDUCP_A` use 97/98/99 for missing/non-response, while
`MEDDL12M_A` and `NOTCOV_A` use 7/8/9. A single blanket `na.strings`
won't work here because `EDUCP_A` also uses 8 and 9 as **valid**
response categories (Bachelor's and Master's degree). So each variable
is cleaned individually against its own codebook.

```{r clean-missing}
adult2024_clean <- adult2024 %>%
  mutate(
    HISPALLP_A = ifelse(HISPALLP_A %in% c(97, 98, 99), NA, HISPALLP_A),
    EDUCP_A    = ifelse(EDUCP_A %in% c(97, 98, 99), NA, EDUCP_A),
    MEDDL12M_A = ifelse(MEDDL12M_A %in% c(7, 8, 9), NA, MEDDL12M_A),
    NOTCOV_A   = ifelse(NOTCOV_A %in% c(7, 8, 9), NA, NOTCOV_A)
  ) %>%
  filter(!is.na(HISPALLP_A), !is.na(EDUCP_A),
         !is.na(MEDDL12M_A), !is.na(NOTCOV_A))

n_clean <- nrow(adult2024_clean)
n_clean
```


## Recoding and Restricting Categories

Race/ethnicity follows the NHIS `HISPALLP_A` codebook
(1 = Hispanic, 2 = Non-Hispanic White, 3 = Non-Hispanic Black/African
American, 4 = Non-Hispanic Asian); smaller multi-race and AIAN
categories (codes 5--7) are excluded. Education is restricted to High
School, Bachelor's, Master's, and Doctoral degree holders. Rows with
missing data on any predictor of interest or the outcome are dropped
(complete-case analysis).

```{r recode}
adult2024_recoded <- adult2024_clean %>%
  mutate(
    education = case_when(
      EDUCP_A == 4  ~ "High School Degree",
      EDUCP_A == 8  ~ "Bachelors Degree",
      EDUCP_A == 9  ~ "Masters Degree",
      EDUCP_A == 10 ~ "Doctoral Degree",
      TRUE ~ NA_character_
    ),
    insurance = case_when(
      NOTCOV_A == 1 ~ "Not covered",
      NOTCOV_A == 2 ~ "Covered",
      TRUE ~ NA_character_
    ),
    race_ethnicity = case_when(
      HISPALLP_A == 1 ~ "Hispanic",
      HISPALLP_A == 2 ~ "White",
      HISPALLP_A == 3 ~ "Black/African American",
      HISPALLP_A == 4 ~ "Asian",
      TRUE ~ NA_character_
    ),
    delayed_dummy = case_when(
      MEDDL12M_A == 1 ~ 1,
      MEDDL12M_A == 2 ~ 0,
      TRUE ~ NA_real_
    ),
    income_level = case_when(
      RATCAT_A %in% c(1, 2, 3)     ~ "Below poverty (<100%)",
      RATCAT_A %in% c(4, 5, 6, 7)  ~ "Near poverty (100-199%)",
      RATCAT_A %in% c(8, 9, 10, 11) ~ "Middle income (200-399%)",
      RATCAT_A %in% c(12, 13, 14)  ~ "Higher income (400%+)",
      TRUE ~ NA_character_
    ),
    income_level = factor(income_level,
      levels = c("Below poverty (<100%)", "Near poverty (100-199%)",
                 "Middle income (200-399%)", "Higher income (400%+)"))
  )

df_complete <- adult2024_recoded %>%
  filter(
    !is.na(education),
    !is.na(insurance),
    !is.na(race_ethnicity),
    !is.na(delayed_dummy)
  ) %>%
  mutate(
    education = factor(education,
      levels = c("High School Degree", "Bachelors Degree",
                 "Masters Degree", "Doctoral Degree")),
    insurance = factor(insurance, levels = c("Covered", "Not covered")),
    race_ethnicity = factor(race_ethnicity,
      levels = c("White", "Black/African American", "Asian", "Hispanic"))
  )

n_final <- nrow(df_complete)
n_final
```

The analytic sample contains 19303 adults.
The reduction reflects the deliberate category restrictions above (not solely item nonresponse): adults outside the four education and four race/ethnicity groups are
excluded by design.

## Descriptive Analysis

### Table 1: Adults by Education

```{r table1}
df_complete %>%
  count(education, name = "Count") %>%
  mutate(Percent = scales::percent(Count / sum(Count), accuracy = 0.1)) %>%
  rename(Education = education) %>%
  as.data.frame()
```

### Table 2: Adults by Race/Ethnicity

```{r table2}
df_complete %>%
  count(race_ethnicity, name = "Count") %>%
  mutate(Percent = scales::percent(Count / sum(Count), accuracy = 0.1)) %>%
  rename(`Race/Ethnicity` = race_ethnicity) %>%
  as.data.frame()
```

### Table 3: Adults by Insurance Status

```{r table3}
df_complete %>%
  count(insurance, name = "Count") %>%
  mutate(Percent = scales::percent(Count / sum(Count), accuracy = 0.1)) %>%
  rename(`Insurance Status` = insurance) %>%
  as.data.frame()
```

### Distribution Plots

```{r plot-education}
ggplot(df_complete, aes(x = education)) +
  geom_bar(fill = "#4C72B0") +
  labs(title = "Distribution of Education Level",
       x = "Education", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r plot-insurance}
ggplot(df_complete, aes(x = insurance)) +
  geom_bar(fill = "#55A868") +
  labs(title = "Distribution of Insurance Coverage",
       x = "Insurance Status", y = "Count") +
  theme_minimal()
```

```{r plot-race}
ggplot(df_complete, aes(x = race_ethnicity)) +
  geom_bar(fill = "#C44E52") +
  labs(title = "Distribution of Race/Ethnicity",
       x = "Race/Ethnicity", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r plot-income}
ggplot(df_complete, aes(x = income_level)) +
  geom_bar(fill = "#8172B2") +
  labs(title = "Distribution of Income Level (Ratio of Family Income to Poverty Threshold)",
       x = "Income Level", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

Income is summarized here for descriptive context; it is not included in
the regression models below, which focus on the three hypothesized
predictors. Adding income as a covariate is a natural extension.

## Proportions of Delayed Care by Group

### Table 4: Delayed Care by Insurance Status

```{r table4}
df_complete %>%
  group_by(insurance) %>%
  summarize(n = n(), prop_delayed = mean(delayed_dummy)) %>%
  mutate(prop_delayed = scales::percent(prop_delayed, accuracy = 0.1)) %>%
  rename(`Insurance Status` = insurance, N = n,
         `% Delayed Care` = prop_delayed) %>%
  as.data.frame()
```

### Table 5: Delayed Care by Education

```{r table5}
df_complete %>%
  group_by(education) %>%
  summarize(n = n(), prop_delayed = mean(delayed_dummy)) %>%
  mutate(prop_delayed = scales::percent(prop_delayed, accuracy = 0.1)) %>%
  rename(Education = education, N = n,
         `% Delayed Care` = prop_delayed) %>%
  as.data.frame()
```

### Table 6: Delayed Care by Race/Ethnicity and Insurance Status

```{r table6}
df_complete %>%
  group_by(race_ethnicity, insurance) %>%
  summarize(n = n(),
            prop_delayed = mean(delayed_dummy),
            .groups = "drop") %>%
  mutate(prop_delayed = scales::percent(prop_delayed, accuracy = 0.1)) %>%
  rename(`Race/Ethnicity` = race_ethnicity,
         `Insurance Status` = insurance, N = n,
         `% Delayed Care` = prop_delayed) %>%
  as.data.frame()
```

```{r fig-race-insurance}
df_complete %>%
  group_by(race_ethnicity, insurance) %>%
  summarize(prop_delayed = mean(delayed_dummy), .groups = "drop") %>%
  ggplot(aes(x = race_ethnicity, y = prop_delayed, fill = insurance)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Delayed Medical Care by Race/Ethnicity and Insurance Status",
    x = NULL,
    y = "Proportion Delayed Care",
    fill = "Insurance Status"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

```{r fig-education-insurance}
df_complete %>%
  group_by(education, insurance) %>%
  summarize(prop_delayed = mean(delayed_dummy), .groups = "drop") %>%
  ggplot(aes(x = education, y = prop_delayed, fill = insurance)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Delayed Medical Care by Education and Insurance Status",
    x = NULL,
    y = "Proportion Delayed Care",
    fill = "Insurance Status"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

**Descriptive takeaways.** Differences in delayed care across
race/ethnicity groups are modest among insured adults but widen
considerably among the uninsured, suggesting insurance coverage — not
race/ethnicity itself — may drive much of the raw disparity. Cell sizes
matter when reading the uninsured bars: several race-by-insurance cells
contain fewer than 150 respondents (see Table 6), so those proportions
carry wide uncertainty and should be interpreted cautiously. The formal
models below test these patterns directly.

## Likelihood Ratio Tests: Bivariate Models vs. Null Model

Each predictor is first tested on its own against an intercept-only
model.

```{r lrt-models}
null_model <- glm(delayed_dummy ~ 1, family = "binomial",
                  data = df_complete)

model_education_biv <- glm(delayed_dummy ~ education,
                           family = "binomial", data = df_complete)
model_race_biv      <- glm(delayed_dummy ~ race_ethnicity,
                           family = "binomial", data = df_complete)
model_insurance_biv <- glm(delayed_dummy ~ insurance,
                           family = "binomial", data = df_complete)

lr_row <- function(model, label) {
  a <- anova(null_model, model, test = "Chisq")
  tibble(
    Predictor = label,
    `Chi-sq`  = a$Deviance[2],
    df        = a$Df[2],
    `p-value` = format.pval(a$`Pr(>Chi)`[2], digits = 3, eps = 0.001)
  )
}

lrt_table <- bind_rows(
  lr_row(model_education_biv, "Education"),
  lr_row(model_race_biv, "Race/Ethnicity"),
  lr_row(model_insurance_biv, "Insurance")
)

# Likelihood ratio tests of each bivariate model against the null model
lrt_table %>%
  mutate(`Chi-sq` = round(`Chi-sq`, 2)) %>%
  as.data.frame()
```

```{r lrt-values, include=FALSE}
# Pull values for inline reporting
lr_stat <- function(model) {
  a <- anova(null_model, model, test = "Chisq")
  p <- a$`Pr(>Chi)`[2]
  list(chisq = a$Deviance[2],
       p = if (p < 0.001) "p < .001" else sprintf("p = %.3f", p))
}
lr_edu <- lr_stat(model_education_biv)
lr_race <- lr_stat(model_race_biv)
lr_ins <- lr_stat(model_insurance_biv)
```

## Multivariable Logistic Regression

```{r multivariate-model}
model_multivariate <- glm(
  delayed_dummy ~ insurance + race_ethnicity + education,
  family = "binomial",
  data = df_complete
)

lr_full <- lr_stat(model_multivariate)
anova(null_model, model_multivariate, test = "Chisq")
```

```{r or-table}
term_labels <- c(
  "insuranceNot covered" = "Uninsured (vs. Covered)",
  "race_ethnicityBlack/African American" =
    "Black/African American (vs. White)",
  "race_ethnicityAsian" = "Asian (vs. White)",
  "race_ethnicityHispanic" = "Hispanic (vs. White)",
  "educationBachelors Degree" = "Bachelor's (vs. High School)",
  "educationMasters Degree" = "Master's (vs. High School)",
  "educationDoctoral Degree" = "Doctoral (vs. High School)"
)

or_table <- tidy(model_multivariate, exponentiate = TRUE,
                 conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(Predictor = recode(term, !!!term_labels))

or_table %>%
  transmute(
    Predictor,
    `Odds Ratio` = sprintf("%.2f", estimate),
    `95% CI` = sprintf("[%.2f, %.2f]", conf.low, conf.high),
    `p-value` = format.pval(p.value, digits = 3, eps = 0.001)
  ) %>%
  as.data.frame()
```

```{r inline-helpers, include=FALSE}
# Helpers so every number reported in the text is computed from the
# fitted model — the prose can never drift out of sync with the output.
get_row <- function(t) filter(or_table, term == t)
fmt_or  <- function(t) sprintf("%.2f", get_row(t)$estimate)
fmt_ci  <- function(t) sprintf("[%.2f, %.2f]",
                               get_row(t)$conf.low, get_row(t)$conf.high)
fmt_p   <- function(t) {
  p <- get_row(t)$p.value
  if (p < 0.001) "p < .001" else sprintf("p = %.3f", p)
}
pct_lower <- function(t) sprintf("%.0f%%", (1 - get_row(t)$estimate) * 100)
```

## Odds Ratio Forest Plot

```{r or-forest-plot}
ggplot(or_table,
       aes(x = reorder(Predictor, estimate), y = estimate)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high),
                  color = "#2166AC", size = 0.7) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  scale_y_log10() +
  coord_flip() +
  labs(
    title = "Odds Ratios with 95% CIs from Multivariable Logistic Regression",
    subtitle = "Outcome: Delayed medical treatment due to cost",
    x = NULL,
    y = "Odds Ratio (log scale)"
  ) +
  theme_minimal()
```

## Interpretation and Discussion

**Bivariate models.** All three predictors significantly improved model
fit over the null model. Education showed the weakest association
($\chi^2$ = 18.66, p < .001), race/ethnicity was stronger
($\chi^2$ = 48.97, p < .001), and insurance coverage showed the
strongest association by far ($\chi^2$ = 472.23, p < .001).

**Multivariable model.** The full model with all three predictors
significantly improved fit over the null model
($\chi^2$ = 527.73, p < .001).

- **Insurance** was the strongest predictor: uninsured adults had about
  6.55 times the odds of delaying care compared to insured adults
  (95% CI [5.60, 7.65], p < .001), even after controlling for
  race/ethnicity and education.
- **Race/ethnicity**: after adjustment, Black/African American adults
  (OR = 0.91, p = 0.348) and Hispanic adults (OR = 0.91, p = 0.296)
  did not differ significantly from White adults. Asian adults were
  the exception, with about 55% lower odds than White adults
  (OR = 0.45, p < .001).
- **Education**: contrary to the original hypothesis, Bachelor's degree
  holders had higher odds of delaying care than high school graduates
  (OR = 1.21, p = 0.005), as did Master's degree holders (OR = 1.29,
  p = 0.002). Doctoral degree holders trended lower (OR = 0.76,
  p = 0.067).
**Takeaway.** Raw racial disparities in delayed care appear to be driven
largely by unequal insurance access rather than race/ethnicity itself:
group differences attenuate substantially once insurance and education
are controlled for. Insurance coverage remains the most direct policy
lever for reducing cost-related care delays.

## Limitations

- **Unweighted estimates.** This analysis does not apply NHIS survey
  weights (`WTFA_A`) or account for the complex sampling design, so
  estimates describe the analytic sample rather than the U.S. adult
  population. A design-based extension using the `survey` package would
  produce population-representative estimates.
- **Complete-case analysis.** Listwise deletion and deliberate category
  restrictions reduce the sample and could introduce selection effects
  if missingness is not random.
- **Cross-sectional data.** Associations are not causal; unmeasured
  confounders such as income, health status, and geography likely
  influence both coverage and care-seeking. Income in particular is a
  strong candidate covariate for future model iterations.
